Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-22 paper.tex
Update narrative for mutation generation

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}

@@

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format. Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
The output of the combined simulations of ancestry and mutations is
then some representation of the resulting nucleotide sequences
(or perhaps some derived statistics of those sequences);
information about the mutations themselves is not typically
provided to the user.

The approach taken in \msprime\ and \tskit\ is quite different.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
integral part of the tree sequence data model, and genetic
variation is encoded by sites and the mutations at those sites.
This may seem at first inconvenient, but is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we required $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space~\citep{kelleher2016efficient}. For
large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster for large
sample sizes~\citep{ralph2020efficiently}.

Motivated by the large numbers of SARS-CoV-2 genomes currently
available, \cite{turakhia2021ultrafast} propose a similar approach
to encoding variation data for non-recombining organisms,
where a Newick tree is annotated by mutations~\citep{mcbroome2021daily}.
Similarly, the Phastsim~\citep{demaio2021phastsim} mutation simulator,
developed primarily to support simulation of SARS-CoV-2 trees
with hundreds of thousands of leaves, can optionally output a
Newick tree, where nodes are annotated with the mutations.
However, the dependence on the Newick format for encoding
topological information means that it suffers from all the
problems of newick [what?], and there is no software library support
[confirm].

\jkcomment{stuff to cover somewhere in this section}
\begin{itemize}
\item Mutation simulations is only included in msprime for historical reasons.
It would be better to see msprime's mutation simulation as a separate
package. The mutation and ancestry code in msprime have no priveledged
access to each other, and mutation simulation would be no less efficient
if done in a different package.
\item Really emphasise this point: you can use msprime's mutation simulation
on \emph{any} tskit tree sequence, produced by SLiM, fwdpp, etc or
even tsinfer. It's completely modular.
\end{itemize}

Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.
The results of mutation simulations across a range of models are
extensively validated
against both theoretical expectations and output from
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

@@

Fig.~\ref{fig-mutations-perf} shows that the mutation generation
machinery in \msprime\ is extremely efficient, requiring
fractions of a second to generate mutations for most simulations.
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
Generating mutations for a tree sequence is particularly efficient
because we can work edge-by-edge rather than having to
consider each tree independently.

@@

Another substantial advantage of our approach is that the simulated
mutations are stored directly in the output data format with
a defined API. Although
Pyvolve and Phastsim are implemented in Python, neither provide
programatic access to the simulated output, outputting instead
to a series of files which must subsequently be parsed.
See the Data Model section for more details and discussion.

@@

\subsection*{Simulation interface}

>>>>

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\tskit}[0]{\texttt{tskit}}
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\scrm}[0]{\texttt{scrm}}
\newcommand{\SLiM}[0]{\texttt{SLiM}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\SeqGen}[0]{\texttt{Seq-Gen}}
\newcommand{\Pyvolve}[0]{\texttt{Pyvolve}}
\newcommand{\phastsim}[0]{\texttt{phastsim}}

@@

\subsection*{Simulating mutations}
\label{sec-mutations}
% Explain the background: what have people done classically?
Because coalescent simulations are usually concerned with
neutral evolution (see the \nameref{sec-selection} section, however)
the problem of generating synthetic genetic variation can be decomposed into
two independent steps:
simulating genetic ancestry (the trees), and then simulating
variation by superimposing mutation processes on those trees.
The classical \texttt{Seq-Gen} program~\citep{rambaut1997seq}
allows a range of different models of sequence evolution to be imposed on a given
tree in Newick format. Population genetic simulators have tended to
include their own implementations of mutation simulation, with
most just supporting the infinite sites
model~\citep[e.g.][]{hudson2002generating}
but with several supporting a range of different models
of sequence
evolution~\citep[e.g.][]{mailund2005coasim,excoffier2011fastsimcoal,
virgoulay2021gspace}.
% Mutations are not treated as first-class data --- the sequences
% were really all people were after, classically.
The output of the combined simulations of ancestry and mutations is
then some representation of the resulting nucleotide sequences
(or some derived statistics of those sequences);
information about the mutations that have been simulated
is usually not retained or made available for subsequent analysis.

The approach taken in \msprime\ is quite different.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are an
integral part of \tskit's tree sequence data model, and genetic
variation is encoded by sites and the mutations at those sites;
the sequences themselves are never stored.
This may seem inconvenient at first, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we required $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space~\citep{kelleher2016efficient},
assuming we have $O(1)$ mutations at each site. For
large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

The advantages of storing mutations rather than sequences have
recently become clear in the context of the vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format---consisting of a Newick tree
and associated mutations in a binary format---and
associated \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format, however) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustate the advantages of the general
approach of storing ancestry and mutations over sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.

% on the other hand, tskit is awesome.
The \tskit\ library provides first-class support for site and mutation
data, fully integrated with the topological information encoded in the
node and edge tables (Fig~\ref{fig-ts-example}). The mature and
efficient C and Python interfaces makes it straightforward to add
further information to an existing file, and to output the results.
As such, it is essentially a historical accident
that \msprime\ (whose main focus is on ancestry simulation)
contains code to simulate mutations: the process could (and arguably should)
be performed in a separate package, and there would be no performance
penalty for doing so. It is important to note that \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real data
using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
or \texttt{Relate}~\citep{speidel2019method}. It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

For version 1.0 we have greatly extended the sophistication of
\msprime's mutation generation engine, achieving near feature-parity
with \SeqGen.
We  support a large number of mutation models, including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
The new \texttt{sim\_mutations} function
(see the \nameref{sec-sim-interface} section) defaults to a finite-sites
mutation model, but full support for the
original infinite sites model is also maintained.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Generating simulations in \msprime\ is extremely efficient.
Fig.~\ref{fig-mutations-perf} shows the time requried to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with $508,125$ edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge (see Fig.\ref{fig-ts-example}) basis, rather than
tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider $273,887,838$ branches
since there are $137,081$ trees and $1,998$ branches in each
tree ($\sim500$ times more work).

Even when we consider a single tree, \msprime\ is far more
efficient than methods such as \SeqGen\ and \Pyvolve,
and has similar performance to \phastsim.

@@

@@

@@

\subsection*{Simulation interface}
\label{sec-sim-interface}
